\bf Looking at Non-Linear Dimension Reductions as Models in the Data Space

### 2024-04-05 paper.qmd
Add first draft at notation section

At first glance, thinking of NLDR as a model fitted to the data might seem strange. It is a model in the sense that it is a "a simplified representation or abstraction of a system, process, or phenomenon in the real world". The \pD observations are the realization of the phenomenon, and the \kD NLDR layout is the simplified representation. From a statistical perspective we can consider the distances between points in the \kD layout to be variance that the model explains, and the (relative) difference with their distances in \pD is the error, or unexplained variance. Abstractly, we can also imagine that the positioning of points in 2D represents fitted values, that will have some prescribed position in \pD that can be compared with the observed value.

<!--### Notation -->

@@

We need to have some math that gives precision to the paragraphs in the previous section, defining mapping from \pD to \kD to get the \kD layout. Might be written below, too (XXX Michael??)

@@

### Generating the \kD layout

The usual procedures are followed to produce a \kD layout, using the preferred method, t-SNE, UMAP, ... The data may need to be pre-processed by standardising or transforming to principal components.

### Constructing the 2D model

>>>>

At first glance, thinking of NLDR as a model fitted to the data might seem strange. It is a model in the sense that it is a "a simplified representation or abstraction of a system, process, or phenomenon in the real world". The \pD observations are the realization of the phenomenon, and the \kD NLDR layout is the simplified representation. From a statistical perspective we can consider the distances between points in the \kD layout to be variance that the model explains, and the (relative) difference with their distances in \pD is the error, or unexplained variance. Abstractly, we can also imagine that the positioning of points in 2D represents fitted values, that will have some prescribed position in \pD that can be compared with the observed value. 
<!--### Notation -->

@@

We define the NLDR model as a function that maps \pD to \kD, and takes some (hyper)parameters $\vec{\theta}$. That is,

$$g : \mathbb{R}^p \rightarrow \mathbb{R}^k.$$

The parameter vector, $\vec{\theta}$, depends on the choice of $g$, and can be considered part of 'model fitting' in the traditional sense. Common choices for $g$ include tSNE, uMAP, PACMAP, or PHATE, although in theory any function that maps the entire space $\mathbb{R}^p$ to a subset of $\mathbb{R}^k$ is suitable. Any input requirements for the data (such as normalisation, or preprocessing through the use of PCA or similar) is considered part of the function $g$.

Next, we create a model of the \kD representation of the data using hexagonal binning. The aim of this process is to reduce the noise in the original data that may be preserved by $g$. Define a hexagonal grid across the \kD space, defined by the height and width of the hexagon ($h_b$ and $w_b$ respectively), the aspect ratio of the hexagon ($h_r$), and the starting coordinates of the hexagonal grid $(x_s, y_s)$. We deliberately separate out the creation of the hexagonal grid from the mapping of points on the grid.

With the grid defined, the \kD points will belong to a hexagonal bin. That is, for each $y \in \mathbfit{Y}$, we can (uniquely) identify the hexagon that the point belongs to. Let,

$$h : \mathbb{R}^k \rightarrow \mathbb{R}^k, $$

be the function that maps a point in \kD space to it's hexagonal centroid. It follows that $h \circ g : \mathbb{R}^p \rightarrow \mathbb{R}^k$ maps each point in $\mathbfit{X}$ to a \kD centroid.

The final step of the algorithm is _lifting_ the \kD model back to \pD, completing the 'model-in-data-space' picture. For this, define the set $H^k$ to be the set of points that belong to centroid $k$. That is, $H^k = \{ y \; | \; g(y) = (h_x^k, h_y^k)\}$. We use the \pD Euclidean mean of the points in $H^k$ to map the centroid $(h_x^k, h_y^k)$ to a point in \pD. Let the $i$-th component of the \pD mean be

$$f_i = \frac{1}{\left|H^k\right|}\sum_{y \in H^k} y_i,$$

with corresponding \pD vector $\vec{f_i}$. Then, $f(h\circ g)(x)$ maps a \pD point to the \pD model estimate, completing the model cycle.

@@

### 2024-04-05 paper.tex
Add first draft at notation section

We need to have some math that gives precision to the paragraphs in the
previous section, defining mapping from \pD to \kD to get the
\kD layout. Might be written below, too (XXX Michael??)

\subsection{\texorpdfstring{Generating the
\kD layout}{Generating the layout}}\label{generating-the-layout}

The usual procedures are followed to produce a \kD layout, using the
preferred method, t-SNE, UMAP, \ldots{} The data may need to be
pre-processed by standardising or transforming to principal components.

\subsection{Constructing the 2D model}\label{constructing-the-2d-model}

>>>>

We define the NLDR model as a function that maps \pD to \kD, and takes
some (hyper)parameters \(\vec{\theta}\). That is,

\[g : \mathbb{R}^p \rightarrow \mathbb{R}^k.\]

The parameter vector, \(\vec{\theta}\), depends on the choice of \(g\),
and can be considered part of `model fitting' in the traditional sense.
Common choices for \(g\) include tSNE, uMAP, PACMAP, or PHATE, although
in theory any function that maps the entire space \(\mathbb{R}^p\) to a
subset of \(\mathbb{R}^k\) is suitable. Any input requirements for the
data (such as normalisation, or preprocessing through the use of PCA or
similar) is considered part of the function \(g\).

Next, we create a model of the \kD representation of the data using
hexagonal binning. The aim of this process is to reduce the noise in the
original data that may be preserved by \(g\). Define a hexagonal grid
across the \kD space, defined by the height and width of the hexagon
(\(h_b\) and \(w_b\) respectively), the aspect ratio of the hexagon
(\(h_r\)), and the starting coordinates of the hexagonal grid
\((x_s, y_s)\). We deliberately separate out the creation of the
hexagonal grid from the mapping of points on the grid.

With the grid defined, the \kD points will belong to a hexagonal bin.
That is, for each \(y \in \mathbfit{Y}\), we can (uniquely) identify the
hexagon that the point belongs to. Let,

\[h : \mathbb{R}^k \rightarrow \mathbb{R}^k, \]

be the function that maps a point in \kD space to it's hexagonal
centroid. It follows that
\(h \circ g : \mathbb{R}^p \rightarrow \mathbb{R}^k\) maps each point in
\(\mathbfit{X}\) to a \kD centroid.

The final step of the algorithm is \emph{lifting} the \kD model back to
\pD, completing the `model-in-data-space' picture. For this, define the
set \(H^k\) to be the set of points that belong to centroid \(k\). That
is, \(H^k = \{ y \; | \; g(y) = (h_x^k, h_y^k)\}\). We use the
\pD Euclidean mean of the points in \(H^k\) to map the centroid
\((h_x^k, h_y^k)\) to a point in \pD. Let the \(i\)-th component of the
\pD mean be

\[f_i = \frac{1}{\left|H^k\right|}\sum_{y \in H^k} y_i,\]

with corresponding \pD vector \(\vec{f_i}\). Then, \(f(h\circ g)(x)\)
maps a \pD point to the \pD model estimate, completing the model cycle.
